---
title: "Reproducible and portable workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reproducible and portable workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = requireNamespace("allofus", quietly = TRUE) &&
    requireNamespace("duckdb", quietly = TRUE) &&
    requireNamespace("testthat", quietly = TRUE)
)
```

The ["getting started"](mockallofus.html) vignette covers interactive use. This
one covers using `mockallofus` to make All of Us analysis code **reproducible**
and **portable**: building a fixed local database, writing one script that runs
both locally and on the Researcher Workbench, and testing analysis code in CI
without the Workbench.

```{r libs}
library(mockallofus)
library(allofus)
suppressMessages(library(dplyr))
```

## Build a database once, reuse it

`mock_aou_connect()` builds a database on first use --- by default
`mock_aou.duckdb` in your current project directory ([`default_mock_db_path()`]),
so each project keeps its own. For a reproducible project you usually want an
explicit, version-pinned database: control the size and the random seed, build
it to a known path, seed your study's data, and reuse it. Generation is
deterministic, so the same `seed` always yields the same database.

```{r build}
db_path <- tempfile(fileext = ".duckdb") # for a real project, a stable path

build_mock_db(db_path, n_persons = 500, seed = 42, quiet = TRUE)

con <- mock_aou_connect(db_path, quiet = TRUE)
mock_seed_concept_set(con, c(201826, 4193704), domain = "condition",
                      prevalence = 0.3, seed = 1, quiet = TRUE)
```

Because the database is a file, seeded data persists. You can disconnect and
reconnect later (in another session, another script) and the data is still
there --- so you build and seed once, then iterate on analysis code against a
stable dataset.

```{r persist}
mock_aou_disconnect(con)

con <- mock_aou_connect(db_path, quiet = TRUE) # reopen the same file
tbl(con, "condition_occurrence") |>
  filter(condition_concept_id %in% c(201826, 4193704)) |>
  summarise(people = n_distinct(person_id))
```

`build_mock_db()` does nothing if the file already exists; pass
`overwrite = TRUE` to rebuild. The default location is the project directory
([`default_mock_db_path()`]); override it for all calls with
`options(mockallofus.path = "...")`. A common project setup is a small script
that builds and seeds the database (committed to your repo), with the `.duckdb`
file itself git-ignored and rebuilt on demand.

## Raw SQL with `aou_sql()`

`aou_sql()` works locally as long as the SQL is portable. The `{CDR}` template
resolves to the local schema, just as it resolves to the curated data
repository on the Workbench.

```{r aou-sql}
aou_sql("
  SELECT gender_concept_id, COUNT(*) AS n
  FROM `{CDR}.person`
  GROUP BY gender_concept_id
", collect = TRUE)
```

BigQuery-only SQL functions (`COUNTIF`, `APPROX_QUANTILES`, ...) will not run on
the local backend; a couple of common ones (`CONTAINS_SUBSTR`,
`REGEXP_CONTAINS`) are emulated. Keep raw SQL portable, or guard
Workbench-specific queries (see the next section). dplyr/dbplyr code, which is
translated per-backend, is portable automatically.

## One script, both places

`allofus::on_workbench()` reports whether you are on the Researcher Workbench.
Route your connection through it and the *same script* connects to the real data
on the Workbench and to the mock database locally:

```{r connect-helper}
connect_aou <- function() {
  if (on_workbench()) {
    aou_connect()          # real All of Us data (BigQuery)
  } else {
    mock_aou_connect()     # local synthetic data (DuckDB)
  }
}
```

Write your analysis as a function of the connection (defaulting to the stored
option, exactly as the `allofus` functions do). Nothing inside it is
mock-specific:

```{r analysis-fn}
high_a1c_cohort <- function(threshold = 6.5, con = getOption("aou.default.con")) {
  aou_concept_set(concepts = c(3004410, 3005673), domains = "measurement",
                  output = "all", concept_set_name = "a1c", con = con) |>
    filter(value_as_number > threshold) |>
    distinct(person_id)
}
```

Locally, seed the inputs and run it; on the Workbench, the identical call runs
against real data.

```{r run-analysis}
mock_seed_concept_set(con, c(3004410, 3005673), domain = "measurement",
                      prevalence = 0.4, values = c(5, 11), seed = 3, quiet = TRUE)

high_a1c_cohort(con = con) |> tally()
```

## Test analysis code without the Workbench

Because the mock database is deterministic and local, you can unit-test analysis
functions in CI. A typical `testthat` setup builds and seeds a small database,
then checks your function's behavior against data you control. For example,
seeding A1c values in `c(5, 11)` at a known prevalence lets you assert the
high-A1c cohort is non-empty but not everyone:

```{r test, results = "hide"}
library(testthat)

test_that("high_a1c_cohort selects participants above the threshold", {
  # in a package, this would live in tests/testthat/ with the db built in setup
  path <- withr::local_tempfile(fileext = ".duckdb")
  build_mock_db(path, n_persons = 300, seed = 99, quiet = TRUE)
  con <- mock_aou_connect(path, quiet = TRUE)
  withr::defer(mock_aou_disconnect(con))
  mock_seed_concept_set(con, c(3004410, 3005673), domain = "measurement",
                        prevalence = 0.5, values = c(5, 11), seed = 1, quiet = TRUE)

  result <- high_a1c_cohort(con = con) |> collect()

  expect_s3_class(result, "data.frame")
  expect_true("person_id" %in% names(result))
  expect_gt(nrow(result), 0)
  expect_lt(nrow(result), 300)
})
```

```{r test-confirm, echo = FALSE}
cat("Test passed.")
```

```{r cleanup, include = FALSE}
mock_aou_disconnect(con)
```

This gives you a fast, offline, reproducible test suite for code that ultimately
runs on real All of Us data --- catching regressions before you spend Workbench
time.

## Summary

- Build a deterministic database with `build_mock_db(path, n_persons, seed)` and
  reuse the file across sessions.
- Keep raw `aou_sql()` portable; dplyr/dbplyr is portable automatically.
- Route connections through `on_workbench()` so one script runs in both places.
- Test connection-taking analysis functions against a seeded mock database in CI.
